---
title: "The Opioids Problem Set"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This problem set was designed with the goal of teaching some basic functions, syntax, and workflow in R. The dataset in question is a profile of opioids generated by the DEA and provided by the Washington Post. This exercise will use the tidyverse, tigris, viridis, tidycensus, gganimate, and sf packages. 

```{r}
#this is a code chunk, you can create one yourself with "Insert" -> "R" or by just typing "```{r}" and pressing enter. You can comment with the hash key. 

library(tidyverse)#here's how to load a package. make sure you install it in the console with install_packages(PACKAGE_NAME) first. 
```

1. Import the opioids dataset and the counties shapefile. 
2. What are the names of the county-year-labeler observations that have the ten highest numbers of opioid pills sold? 
3. What are the states with the ten highest total number of opioid pills sold over the years 2006-2012? Which manufacturers sold the ten highest number of opioid pills over the years 2006-2012? 
4. Make a choropleth map representing total number of opioid pills sold over 2006-2012 at the county level. Add the viridis color scale and remove the borders of the counties to make it look nicer. 
5. Notably, this map is disproportionately skewed towards high-population counties (AKA major cities). This would still be useful, but a more common map would be per-capita information on opioid pills. Let’s assume county populations are constant from 2006-2012, the years available with this dataset, an assumption that I’m fairly comfortable with making. Download total county populations from the 2007-2012 5-year ACS using the tidycensus package. 
6. Merge this into the opioids dataset and create a new column representing opioid pills sold per capita. 
7. Make sure you’ve aggregated to total pills per county over 2006-2012 and make a new map, again with the viridis scale and county borders removed. Take a visual check with the map you made in #5 to assure yourself that the per-capita distinction is indeed an important one. 
8. Let’s take it up a notch! Time is an interesting and relevant variable here, but one that often looks weird with maps. One way is to use a facet_wrap to create a different map for each year. Recreate the map from step 8, but facet by year using the facet_wrap() command. 
9. Another, more fun but slightly harder way to represent time with a map is through an animation. Create an animation (frame # and rate to your preference) that represents the total number of opioid pills sold per year in each county on a map. 
